---
title: "Grammar of Graphics in R"
output: html_notebook
---

```{r load_ggplot, echo = F}
library(ggplot2)
```

# Let's get to coding!

We will start with an example where we take `mpg` data and make a scatterplot.

How do we get from this:

## The data

```{r view_data}
head(mpg)
```

to this:

## The plot

```{r preview_plot, echo = F}
mpg_plot <- ggplot(
    data = mpg,
    mapping = aes(x = cty, y = hwy)
  ) +
  geom_point(
    mapping = aes(col = factor(cyl))
  ) +
  geom_abline(
    slope = 1,
    intercept = 0,
    color = 'black',
    linetype = 'dotted'
  ) +
  coord_fixed(
    ratio = 1
  ) +
  scale_color_manual(
    name = '# of Cyl',
    values = c('skyblue','royalblue', 'blue', 'navy')
  ) +
  ggtitle('Miles per gallon') +
  theme_light()

mpg_plot
```

# Let's plot!

Load the ggplot2 library:

```{r load_library_no_eval, eval = F}
library(ggplot2)
```

`ggplot` is the base layer.

```{r step00_base}
ggplot()
```

This is a good place to for us to define our plot's default data.

```{r step01_data}
ggplot(data = mpg)
```

There isn't much being plotted yet... Let's tell our plot about our mapping, where our `x` will be our `cty` and `y` will be our `hwy`.

```{r step02_mapping}
ggplot(
    data = mpg,
    mapping = aes(x = cty, y = hwy)
  )
```

Our plot now has enough information to set some automatic axis limits based on the data.

```{r step03_points}
ggplot(
    data = mpg,
    mapping = aes(x = cty, y = hwy)
  ) +
  geom_point()
```


```{r step04_colors}
ggplot(
    data = mpg,
    mapping = aes(x = cty, y = hwy)
  ) +
  geom_point(
    mapping = aes(col = factor(cyl))
  )
```

At this point, we have the base of the plot.  We can save this to a variable, `mpg_base_plot`, and make additional adjustments.

```{r step05_assign}
mpg_base_plot <- ggplot(
    data = mpg,
    mapping = aes(x = cty, y = hwy)
  ) +
  geom_point(
    mapping = aes(col = factor(cyl))
  )
mpg_base_plot
```


To add the reference line, we add a layer like so:

```{r step06_line}
mpg_base_plot +
  geom_abline(
    slope = 1,
    intercept = 0,
    color = 'black',
    linetype = 'dotted'
  )
```

Note that this geom layer can be added on top of a blank plot.  All `geom_` prefixed functions return an `environment` object that holds onto the information we call it with.

```{r line_layer_example}
ggplot() +
  geom_abline(
    slope = 1,
    intercept = 0,
    color = 'black',
    linetype = 'dotted'
  )
```

Let's continue.  Our plot looks a bit funny -- the grid for the x and y are not 1 to 1.  Let's use a `coord_` function to adjust this:

```{r step07_fixed}
mpg_base_plot +
  geom_abline(
    slope = 1,
    intercept = 0,
    color = 'black',
    linetype = 'dotted'
  ) +
  coord_fixed(ratio = 1)
```

We can clean up the plot like so:

```{r step08_cleanup}
mpg_plot <- mpg_base_plot +
  geom_abline(
    slope = 1,
    intercept = 0,
    color = 'black',
    linetype = 'dotted'
  ) +
  coord_fixed(ratio = 1) +
  scale_color_manual(
    name = '# of Cyl',
    values = c('skyblue','royalblue', 'blue', 'navy')
  ) +
  ggtitle('Miles per gallon') +
  theme_light()
mpg_plot
```

We can further manipulate the plot using things like `facet_`.

```{r facets_example}
mpg_plot +
  facet_grid(.~class)
```


# The magic behind the `+`

We can discover how the `+` is helping us construct these plots by running:

```{r what_the_plus}
methods('+')
```

From the result, we can see that the `+` function is overloaded for objects that match the `gg` namespace.  If we look [at the code](https://github.com/tidyverse/ggplot2/blob/master/R/plot-construction.r#L39) for this function, we learn more about how it works.


# The makings of a pie

### Grammar-based graphics lends flexibility

Back to Leland Wilkinson's point about making a pie -- that a pie chart is really a stacked bar chart that has been transformed with polar coordinates where the y is mapped to the angle, a.k.a. theta.

### We start with a bar chart:

```{r pie00_bar}
diamonds_by_cut_base <- ggplot(data = diamonds, mapping = aes(x = cut, fill = cut))
diamonds_by_cut_base +
  geom_bar()
```

Here, we have a stacked bar chart of diamonds grouped by cut.

```{r pie01_stacked}
diamonds_by_cut_stacked <- diamonds_by_cut_base +
  geom_bar(width = 1, aes(x=factor("")))

diamonds_by_cut_stacked
```

Taking the bar chart, we simply "add" a `coord_polar` to it.

```{r pie02_bullseye}
diamonds_by_cut_stacked +
  coord_polar()  +
  scale_x_discrete("")
```

By default, `coord_polar` maps the `theta` from `x` and the `radius` from `y`.

```{r what_is_a_coord_anyway}
bullseye_coords <- coord_polar()
bullseye_coords$theta
bullseye_coords$r
```

We need to a `coord_polar` where the `theta` maps from `y`:

```{r pie_coords}
pie_coords <- coord_polar(theta = "y")
pie_coords$theta
```

Now, we can transform the stacked bar chart to a pie chart:

```{r pie03_pie}
diamonds_by_cut_stacked +
  pie_coords
```

```{r}
?coord_polar
```


# Let's experiment

We can get the base plots from the ggplot cheatsheets on our computer by sourcing [this R script](https://houstonusers.github.io/intro-to-ggplot2/examples_from_cheatsheet).

```{r eval = F}
source('https://goo.gl/3EQSXt')
```


# A layered grammar is powerful

The full grammar consists of:
  * The base plot with default data and mappings
  * Any number of layers, each with
    * Data
      * plot's data by default
    * Mapping
      * plot's mappings by default
    * Geom
      * Sometimes defined by default by `stat_` if defined
    * Stat
      * Identity by default
      * Sometimes defined by default by `geom_` if defined
    * Position
      * Optional, any adjustments necessary to geom positioning
Additionally, the grammar includes control over the whole plot's
  * Scales
  * Coordinates
  * Facets


# Reading and Sources

[A Layered Grammar of Graphics](http://vita.had.co.nz/papers/layered-grammar.pdf)

[Mastering the Grammar](https://rpubs.com/flowertear/224424)

[Introduction to R Graphics with `ggplot2`](http://tutorials.iq.harvard.edu/R/Rgraphics/Rgraphics.html)

[R for Data Science: Visualizing Data](http://r4ds.had.co.nz/data-visualisation.html#introduction-1)

[Data Camp: Data Visualization with ggplot2](https://www.datacamp.com/courses/data-visualization-with-ggplot2-1)

[Swirl: Exploratory Data Analysis](http://swirlstats.com/scn/eda.html)

[ggplot2 Cheatsheet](https://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf)

[Telling stories with data using the grammar of graphics](https://codewords.recurse.com/issues/six/telling-stories-with-data-using-the-grammar-of-graphics)

[ggplot2 source code](https://github.com/tidyverse/ggplot2/tree/master/R)

[ggplot2: Elegant Graphics for Data Analysis](http://ggplot2.org/book/)

[The Grammar of Graphics](https://www.amazon.com/Grammar-Graphics-Statistics-Computing/dp/0387245448)

[Semiology of Graphics](https://www.amazon.com/Semiology-Graphics-Diagrams-Networks-Maps/dp/1589482611)
